Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2019-10-17 manuscript/manuscript.Rmd
Anne's edits done

Lake datasets that cover longer time periods and broader spatial scales are now becoming available.  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing]. Studying trends across large spatial scales can illustrate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (e.g. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may or may not be present in individual lakes [@cheruvelil2013multi; @lottig2014long].

@@

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.  For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had approximately `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

@@

Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as "site-specific z-scores".

@@

During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

@@

For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].  However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*. Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

>>>>

Lake datasets that cover longer time periods and broader spatial scales are now becoming available.  Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis. These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing]. Studying temporal trends across large spatial scales can illustrate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental]. Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is indicative of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking]. These trophic state models indicate that landscape variables (e.g. ecoregion, elevation, and latitude) are important and that regional trends exist. Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance]. Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may or may not be present in individual lakes [@cheruvelil2013multi; @lottig2014long].

@@

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.  For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had approximately `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.

@@

Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].  The resultant value is commonly referred to as a z-score.  We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time at individual sites (i.e. lakes), z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as "site-specific z-scores".

@@

During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).

@@

For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].  However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*. Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.
